---
title: "SHEN_CHANG_HW4"
author: "ChangShen"
date: "10/28/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##HW4
1.Set path and package

2.use pacman to install and load packages

```{r 1}
setwd("/Users/dianshen/Desktop/SAS/Assignment2/")
if(!"pacman" %in%installed.packages()) install.packages("pacman")
library(pacman)
p_load(dplyr,tidyr,plyr,knitr,readr)
```

## Read files

###read in the 3 files you need

*Demographics* - include participant information: PATNO, Year, Gender, Residency, Major\n

*Assess_original* - orignal assement by different courses in long format :PATNO,QOL,Scale\n

*Assess_original* - make up assement by differen coursed in long format :PATNO,QOL,Scale\n

```{r 2}
Demographics<-read_csv(url("https://raw.githubusercontent.com/diana12333/CourseCode/master/Advanced%20R/Demographics.csv"),col_names = T)
Assess_original <- read_csv(url("https://raw.githubusercontent.com/diana12333/CourseCode/master/Advanced%20R/Assessment_original.csv"),col_names = T)
Assess_makeup<-read_csv(url("https://raw.githubusercontent.com/diana12333/CourseCode/master/Advanced%20R/Assessment_makeup.csv"),col_names = T)
```

## Self defined function

###Short inline functions for concise considerarion

count_na—— ount the total number of NA   \

RowMeans/RowSums/RowMax——rowmeans  ，rowsums,rowmax function with default setting ignoring NAs   \n
```{r 3}
count_na <- function(x) sum(is.na(x))
RowMeans <- function(x) rowMeans(x,na.rm = TRUE)
RowSums <- function(x) rowSums(x,na.rm = TRUE)
RowMax <- function(x) apply(x,1,function(x) max(x,na.rm = TRUE))
```

## Transfom and combine data
```{r 4}
#combind the assesment data(use make up to update the scales in orignal assements)
combined <- rbind(Assess_original,Assess_makeup)
combined$SCALE <- mapvalues(combined$SCALE,from=c("1","2","3","4","5"),to=paste("course",1:5,sep = ""))

# tranfer the assesment to wide format and combined with the mainparticipant information
FINALDATA <- spread(combined[!is.na(combined$QOL),],SCALE,QOL)%>%
            right_join(.,Demographics,by="PATNO")

#calculate the A/B/C score
# with the aid of [count_na] to evaluate how many courses have missing values
FINALDATA<- FINALDATA%>%
            mutate( countna = apply(FINALDATA[,paste("course",1:5,sep = "")], 1, count_na))%>%
            mutate( Ascore = ifelse(countna==0,.7*RowMeans(FINALDATA[,paste("course",1:4,sep = "")])+.3*course5,NA),
                    Bscore = ifelse(countna<=1,.5*RowMeans(FINALDATA[,paste("course",1:3,sep = "")])+
                                      .5*RowMeans(FINALDATA[,paste("course",4:5,sep = "")]),NA),
                    Cscore = ifelse(Residency==2,25+0.5*RowMeans(FINALDATA[,paste("course",1:5,sep = "")]),
                                   (50*countna+RowSums(FINALDATA[,paste("course",1:5,sep = "")]))/5))
#1.caculate the HighestQOL 
#2.only keep the variable need in the final report
#3.round scores to 2 decimals
#4.order by PATNO
FINALDATA<- FINALDATA%>%
           mutate(High = RowMax(FINALDATA[,c("Ascore","Bscore","Cscore")]))%>%
           mutate(HighestQOL = case_when(
                            High==Ascore ~ "A",
                            High==Bscore ~ "B",
                            High==Cscore ~ "C"
                    ))%>%
          select(PATNO, Year, Residency,Major,Ascore,Bscore,Cscore,HighestQOL)%>%
           mutate_if(is.numeric, round, 2)%>%arrange(PATNO)
```

##Create the labels for factor variables,ie.

1. Year:1 = Freshman, 2= Sophomore, 3=Junior, 4=Senior, 5=Senior Plus

2. Residency: 1=In State, 2=Out of State\\

3. Major: 1=Chemistry, 2=Biology, 3=Mathematics, 4=Physics, 5=Psychology, 6=Other

```{r 5}
FINALDATA$Year <- mapvalues(FINALDATA$Year,  from = c("1","2","3","4","5"), to = c("Freshman","Sophomore","Junior","Senior","Senior Plus") )
FINALDATA$Residency <- mapvalues(FINALDATA$Residency,from=c("1","2"),to=c("In State","Out of State"))
FINALDATA$Major <- mapvalues(FINALDATA$Major, from = c("1","2","3","4","5","6"),to = c("Chemistry", "Biology", "Mathematics", "Physics", "Psychology", "Other"))
```



## Final Ressult Output
Deliverable Output
```{r 6}
#1
kable(FINALDATA[1:15,], caption = "the first 15 observations from FINALDATA")

#2
kable(table(FINALDATA$HighestQOL),caption ="A table with the frequency distribution of HighestQOL")


#3
kable(table(FINALDATA$HighestQOL,FINALDATA$Residency),caption = "A table with the frequency distribution of HighestQOL by Residency")
```
